Limitations on the Two-Infall Scenario in the Era of Large Stellar Age Catalogs

### 2025-04-07 src/tex/ms.tex

Finally, panel (c) of Figure \ref{fig:twoinfall-parameters} shows that the time of maximum infall $t_{\rm max}$ (c) strongly affects the overall stellar abundance distribution for values $t_{\rm max}\leq2\,{\rm Gyr}$, but in this case the gas tracks do not produce the characteristic abundance loop. For $t_{\rm max}>2\,{\rm Gyr}$, varying $t_{\rm max}$ results in a minor shift to the mean of the MDF and little change to the [O/Fe] distributions, even though the abundance tracks in [O/Fe]--[Fe/H] space appear very different. The value of $t_{\rm max}$ also slightly adjusts the ISM abundance endpoint, as a longer $t_{\rm max}$ means the chemical evolution ``reset'' from the second infall occurs closer to the present day (see discussion in Section \ref{sec:age-abundance}. We adopt a fiducial value of $t_{\rm max}=4.2\,{\rm Gyr}$ so that the second infall occurs at a lookback time of 9 Gyr, younger than the median age of the thick disk in the APOKASC-3 catalog of $9.14\pm0.05\,{\rm Gyr}$ \citep{pinsonneault_apokasc-3_2025}. This value is generally in line with previous two-infall studies \citep[e.g.,][]{nissen_high-precision_2020,spitoni_galactic_2020,spitoni_apogee_2021}.

@@

\begin{figure*}
    \centering
    \includegraphics{figures/spitoni_comparison.pdf}
    \caption{Chemical abundance tracks and distributions from one-zone models which use our fiducial parameters (solid curves) and the MCMC-derived parameters from \citet[][dashed curves]{spitoni_apogee_2021}. All models use the $y/Z_\odot=1$ yield set. The time markers and marginal panels are the same as in Figure \ref{fig:twoinfall-parameters}. We also include a 2-D heatmap of the APOGEE stellar abundances in each panel for reference. APOGEE stars are restricted to (a) $2\leq R_{\rm gal}<6$ kpc, (b) $6\leq R_{\rm gal}<120$ kpc, and (c) $10\leq R_{\rm gal}<14$ kpc, and $0\leq |z|<2$ kpc in each panel. \todo{Should I instead compare to Palla et al. (2020), who have a more cohesive multi-zone model rather than three independent one-zone models?}}
    \label{fig:spitoni-comparison}
\end{figure*}

@@

The dilution effect discussed in Section \ref{sec:yields} is clearly seen in the multi-zone model results. Figure \ref{fig:abundance-evolution-yields} shows stellar age--abundance relations produced by models with \yZ{1} and \yZ{2} with fiducial parameters (Table \ref{tab:multizone-parameters}). The \yZ{1} model (column a) shows two clear discrepancies with the \citet{leung_variational_2023} age--abundance relation: a major dilution of $\sim0.5$ dex at a lookback time of $\sim9\,{\rm Gyr}$ near where the data show a maximum in [O/H], and non-zero abundance evolution at late times where the data show very little abundance evolution. The evolution of [Fe/H] is similar, but the approach to Solar abundance is slower because of the additional delay imposed on Fe production from SNe Ia. The \yZ{2} yield set (column b) mitigates both of these issues by shortening the time it takes the ISM metallicity to rebound post-$t_{\rm max}$, producing a much flatter abundance curve at late times. However, model (b) produces a poorer fit to the age--[O/Fe] relation: the decline in [O/Fe] over the thin disk epoch is much steeper than the data.

@@

Figure \ref{fig:abundance-evolution-params} shows the effect of varying several parameters on the \yZ{1} model.

The observed rise in the median abundance of stars with ages of $\sim4-8\,{\rm Gyr}$ is thought to be due to radial migration, as those stars are more likely to have migrated from the dense inner metal-rich regions of the Galaxy during that time. Although our fiducial model does include a prescription for radial migration, the majority of stars in that age range in Figure \ref{fig:abundance-evolution-yields} still have sub-Solar abundances. Column (b) of Figure \ref{fig:abundance-evolution-params} presents a model with $y/Z_\odot=1$ but with a stronger migration prescription of $\sigma_{\rm RM8}=5\,{\rm kpc}$. Here, the stars which make up the present-day Solar neighborhood are drawn from a wider range of birth $R_{\rm gal}$, producing a broader abundance distribution for any given age. However, even though this prescription is much stronger than the estimates of, e.g., \citet{frankel_measuring_2018}, the model cannot reproduce the observed age--abundance trends in [O/H] and [Fe/H].

Second, as in Section \ref{sec:abundance-distributions}, we investigate a model where the infalling gas is enriched to a metallicity $\mathOH=\mathFeH={\rm [X/H]}_{\rm CGM}$ before accreting onto the disk. Column (c) of Figure \ref{fig:abundance-evolution-params} shows results for the case where ${\rm [X/H]}_{\rm CGM}-0.5$, the highest metallicity allowed by the local low-$\alpha$ population. Pre-enriched infall at this level mitigates but does not completely solve the two discrepancies. The dilution effect of the second infall is reduced to the $\sim0.3$-dex level as the gas which replenishes the Galaxy's reservoir is no longer pristine; however, the width of the stellar abundance distribution at any given age is also reduced, because the enriched gas accretion imposes a lower limit on the metallicity of the outermost regions, from which the stars of the low-metallicity tail in the Solar neighborhood are drawn. The late-time gas abundance evolution is similar to the fiducial model, but it ends at slightly super-Solar metallicity --- an effect which can be compensated by a slightly increased value of $\eta$.

Finally, we explore a model where the local thick-to-thin disk surface density ratio is $\sim4$ times larger, $f_\Sigma(R_\odot)=0.48$. This is higher than most of the constraints from population counts or GCE models (see Section \ref{sec:sfh}). Column (d) of Figure \ref{fig:abundance-evolution-params} shows that requiring a more massive thick disk can reduce the dilution and recent evolution of the ISM, similar to the pre-enriched infall, because more of the gas disk is built up during the first infall phase. The model produces the best agreement with the observed age--[Fe/H] relation (second row). However, agreement with the observed age--[O/Fe] relation is poor, as the [O/Fe] of the old low-$\alpha$ disk is $\sim0.1$ dex lower than observed.

Overall, no modification to the \yZ{1} model is able to completely overcome both the dilution and late-time evolution issues.

>>>>

Finally, panel (c) of Figure \ref{fig:twoinfall-parameters} shows that the time of maximum infall $t_{\rm max}$ (c) strongly affects the overall stellar abundance distribution for values $t_{\rm max}\leq2\,{\rm Gyr}$, but in this case the gas tracks do not produce the characteristic abundance loop. For $t_{\rm max}>2\,{\rm Gyr}$, varying $t_{\rm max}$ results in a minor shift to the mean of the MDF and little change to the [O/Fe] distributions, even though the abundance tracks in [O/Fe]--[Fe/H] space appear very different. The value of $t_{\rm max}$ also slightly adjusts the ISM abundance endpoint, as a longer $t_{\rm max}$ means the chemical evolution ``reset'' from the second infall occurs closer to the present day (see discussion in Section \ref{sec:age-abundance}. We adopt a fiducial value of $t_{\rm max}=4.2\,{\rm Gyr}$ so that the second infall occurs at a lookback time of 9 Gyr, younger than the median age of the thick disk in the APOKASC-3 catalog of $9.14\pm0.05\,{\rm Gyr}$ \citep{pinsonneault_apokasc-3_2025}. This value is generally in line with previous two-infall studies \citep[e.g.,][]{nissen_high-precision_2020,spitoni_galactic_2020,spitoni_apogee_2021}. \todo{How does this compare to the age of GSE?}

@@

\begin{figure*}
    \centering
    \includegraphics{figures/spitoni_comparison.pdf}
    \caption{Chemical abundance tracks and distributions from one-zone models which use our fiducial parameters (solid curves) and the MCMC-derived parameters from \citet[][dashed curves]{spitoni_apogee_2021} at three different galactocentric radii. All models use the $y/Z_\odot=1$ yield set and corresponding outflow prescription (see Table \ref{tab:yields}). The time markers and marginal panels are the same as in Figure \ref{fig:twoinfall-parameters}. We also include a 2-D heatmap of the APOGEE stellar abundances in each panel for reference. APOGEE stars are restricted to (a) $2\leq R_{\rm gal}<6$ kpc, (b) $6\leq R_{\rm gal}<120$ kpc, and (c) $10\leq R_{\rm gal}<14$ kpc, and $0\leq |z|<2$ kpc in each panel.}
    \label{fig:spitoni-comparison}
\end{figure*}

@@

The dilution effect discussed in Section \ref{sec:yields} is clearly seen in the multi-zone model results. Figure \ref{fig:abundance-evolution-yields} shows stellar age--abundance relations produced by models with \yZ{1} and \yZ{2} with fiducial parameters (Table \ref{tab:multizone-parameters}). The \yZ{1} model (column a) shows two clear discrepancies with the \citet{leung_variational_2023} age--abundance relation: a major dilution of $\sim0.5$ dex at a lookback time of $\sim9\,{\rm Gyr}$ near where the data show a maximum in [O/H], and non-zero abundance evolution at late times where the data show very little abundance evolution. The evolution of [Fe/H] is similar, but the approach to the final metallicity is slower because of the additional delay imposed on Fe production from SNe Ia. The \yZ{2} yield set (column b) mitigates both of these issues by shortening the time it takes the ISM metallicity to rebound post-$t_{\rm max}$, producing a much flatter abundance curve at late times. However, model (b) produces a poorer fit to the age--[O/Fe] relation: the decline in [O/Fe] over the thin disk epoch is steeper than the data, especially for ages $\sim4-8\,{\rm Gyr}$.

@@

We attempt to mitigate these discrepancies in the age--metallicity relation for the empirical yield scale. 
% Figure \ref{fig:abundance-evolution-params} shows the effect of varying several parameters on the \yZ{1} model. In column (b), we approximately double the strength of radial migration to $\sigma_{\rm RM8}=5.0\,{\rm kpc}$.
For instance, the observed rise in the median metallicity of stars with ages of $\sim4-10\,{\rm Gyr}$ is thought to be due to radial migration, as those stars are more likely to have migrated from the dense inner metal-rich regions of the Galaxy during that time. Although our fiducial model does include a prescription for radial migration, the majority of stars in that age range in Figure \ref{fig:abundance-evolution-yields} still have sub-Solar abundances. Therefore, column (b) of Figure \ref{fig:abundance-evolution-params} presents a model with $y/Z_\odot=1$ and a stronger migration prescription of $\sigma_{\rm RM8}=5\,{\rm kpc}$. As a result, the stars which make up the present-day Solar neighborhood are drawn from a wider range of birth $R_{\rm gal}$, producing a broader abundance distribution for any given age. However, even though this prescription is much stronger than the estimates of, e.g., \citet{frankel_measuring_2018}, the model still significantly under-predicts the metallicity of $\sim4-9\,{\rm Gyr}$ old stars.

Next, as in Section \ref{sec:abundance-distributions}, we investigate a model where the infalling gas is enriched to a metallicity $\mathOH=\mathFeH={\rm [X/H]}_{\rm CGM}$ before accreting onto the disk. Column (c) of Figure \ref{fig:abundance-evolution-params} shows results for the case where ${\rm [X/H]}_{\rm CGM}-0.5$, the highest metallicity allowed by the local low-$\alpha$ population. Pre-enriched infall at this level mitigates but does not completely solve the two discrepancies. The dilution effect of the second infall is reduced to the $\sim0.3$-dex level as the gas which replenishes the Galaxy's reservoir is no longer pristine; however, the width of the stellar abundance distribution at any given age is also reduced, because the enriched gas accretion imposes a lower limit on the metallicity of the outermost regions, from which the stars of the low-metallicity tail in the Solar neighborhood are drawn. The late-time gas abundance evolution is similar to the fiducial model, but it ends at slightly super-Solar metallicity --- an effect which can be compensated by a slightly increased value of $\eta$. This model also narrows the [O/Fe] distribution of mono-age populations (almost all the model stars fall within the $1\sigma$ band of the data), which could be compensated for by stronger radial migration.

Finally, we explore a model where the local thick-to-thin disk surface density ratio is $\sim4$ times larger, $f_\Sigma(R_\odot)=0.48$. This is higher than most of the constraints from population counts or GCE models (see Section \ref{sec:sfh}). Column (d) of Figure \ref{fig:abundance-evolution-params} shows that requiring a more massive thick disk can reduce the dilution and recent evolution of the ISM, similar to the pre-enriched infall, because more of the gas disk is built up during the first infall phase. The model produces the best agreement with the observed age--[Fe/H] relation (second row). However, agreement with the observed age--[O/Fe] relation is poor, with the model predicting less evolution in [O/Fe] over the past $\sim9$ Gyr than observed by $\approx0.1$ dex.

Overall, no modification to the \yZ{1} model is able to completely overcome both the dilution and late-time evolution issues. Pre-enrichment of the accreted gas and a higher disk mass ratio can reduce the discrepancy with the data, but they cause issues of their own in the age--[O/Fe] plane.
